---
title: "pvalues_interactive"
author: "hugo pineda"
date: "11/23/2014"
output: html_document
runtime: shiny
---
This document compare the gene expression values of two groups. For each gene, it perform a t-test with Empirical Bayes correction and multiple testing correction. As inputs, it uses the "Expression.RMA.txt" and the "design.txt" files. 


##Choose the groups
```{r, echo=FALSE, message=FALSE}

#Get the data and load packages
library(limma)
library(reshape2)
library(dplyr)
library(ggplot2)
library(knitr)


#Create the user inputs for selecting the sample groups
selectInput('group1', 'Choose group 1 samples:', choices=c(1:8), multiple=TRUE, selectize=TRUE, selected=c(1,5))
selectInput('group2', 'Choose group 2 samples:', choices=c(1:8), multiple=TRUE, selectize=TRUE, selected=3)

#Read data
design <- read.delim("./design.txt")
design <- filter(design, Dye=="Cy3")
data.rma.full <- read.delim("./Expression.RMA.txt")
annot <- data.rma.full[,c(1:13)]
data.rma <- as.matrix(data.rma.full[,-c(1:13)])
groups <- design$SampleID


#This reactive object contains the user groups choices.
groups.input<-reactive({
  group1 <- as.list(input$group1)
  group2 <- as.list(input$group2)
  
  group.ls<-list("group1"= group1, "group2"=group2)
  
  })

#This reactive object perform the t-test and create a df as output
Pval.annot<-reactive({
  group1 <- as.list(input$group1)
  group2 <- as.list(input$group2)
  
  idx1 <- which(groups %in% group1)
  idx2 <- which(groups %in% group2)
  x <- data.rma[,c(idx1,idx2)]
  
  # The following code will perform a t-test, with Empirical Bayes correction and multiple testing correction:
  factorlevels <- c(rep(1,length(idx1)),rep(2,length(idx2)))
  design.mat <- model.matrix(~ 0+factor(factorlevels))
  colnames(design.mat) <- c("group1", "group2")
  fit <- lmFit(x, design.mat)
  contrast.matrix <- makeContrasts(group2-group1, levels=design.mat)
  fit2 <- contrasts.fit(fit, contrast.matrix)
  fit2 <- eBayes(fit2)
  tblPval <- topTable(fit2, adjust="BH",number=nrow(data.rma))
  
  #Now merge the results with the annot df. It will be useful later to have this information in the same df. 
  #First create a row in both dataframes containing the row name.
  annot$idx<-as.vector(rownames(annot), mode="numeric")
  tblPval$idx<-as.vector(rownames(tblPval), mode="numeric")
  #Merge both df by the new column
  Pval.annot<-merge(annot, tblPval, by="idx")
  #Sort the df in decreasing order by the corrected pvalue 
  Pval.annot<-arrange(Pval.annot, adj.P.Val)
  Pval.annot  
  })


#Create a reactive dataframe containing 4 columns: "GeneName", "mean", "sd", "group".
#"mean" and "sd" contain logratio mean value and standard deviation of each sample. 
#"group" is the combination of "O2"+"time". It is used in the plot
data.plot<-reactive({
  group1 <- as.list(input$group1)
  group2 <- as.list(input$group2)
  #Remove "gFG_" from data.rma.full column names. Later it will be useful to merge with Pval.annot
  colnames(data.rma.full)<-gsub("gFG_", "", colnames(data.rma.full))
  #Remove all annotation columns except the GeneName
  data.rma.full<-data.rma.full[, -c(1:6, 8:13)]
  
  #Get the Array number for each sample in group 1. First get the rows from the samples .
  #I do that to later select the columns of the samples from the data.rmn.full df
  group1df<-filter(design, SampleID %in% group1)
  #get array number as character
  array1<-as.character(unique(group1df$Array))
  #Do the same for group2
  group2df<-filter(design, SampleID %in% group2)
  array2<-as.character(unique(group2df$Array))
  
  #From the input data (data.rma.full), select the columns of the samples and the GeneNames
  genes.wide.df<-data.rma.full[, c("GeneName", array1, array2)]
  
  #Transform from wide to long format, now 3 columns, the geneName, the array number and the expression value
  genes.df<-melt(genes.wide.df)
  #Change the colnames to merge by Array column
  colnames(genes.df)<-c("GeneName", "Array", "logratio")
  #merge with design file by the Array column. 
  #This add append to the genes.df all the information about the samples contained in the design file
  #   design.no.array<-design[,-c(1,4)]
  df<-merge(genes.df, design, by="Array")
  df<-select(df, GeneName, logratio, SampleID, O2, time)
  #Sort by sampleID to plor the results ordered
  DF<-summarise(group_by(df, SampleID, GeneName), mean = mean(logratio), sd=sd(logratio))
  design.m<-unique(select(design, SampleID, O2, time))
  DF<-merge(DF, design.m, by="SampleID")
  DF<-arrange(DF, as.factor(SampleID))
  DF$group<-interaction(DF$time, DF$O2)
  DF
  
  })

```

####Design file:
```{r, echo=FALSE, message=FALSE}
#Print the design file,  only show: Array, SampleID, O2, time columns. 
#Sort by SampleID
renderTable({
  arrange(select(design, Array, SampleID, O2, time), SampleID)
  
  })
```


##Search for a gene
```{r, echo=FALSE, message=FALSE}
# textInput('gene', 'Choose a gene:')
selectInput('gene', 'Choose gene:', choices=as.character(annot$GeneName), multiple=TRUE, selectize=TRUE)


renderPlot({
  gene<-input$gene
  data.plot<-as.data.frame(data.plot())
  data.plot<-filter(data.plot, GeneName %in% gene)
  data.plot<-as.data.frame(data.plot)
  
  p<-ggplot(data.plot, aes(as.factor(SampleID), mean)) + theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.text = element_text(size = 12))
  p + geom_bar(aes(fill=group), stat="identity") +
    #   p + geom_bar(stat="identity") +
    geom_errorbar(aes(ymin= mean - sd, ymax=mean + sd), width=0.25) +
    facet_wrap(~GeneName)
  })

```


###Adjusted P-values of the selected genes
```{r,echo=FALSE, message=FALSE}
renderTable({
  Pval.annot<-Pval.annot()
  genes<-input$gene
  #Avoid to print all the genes with an empty GeneName
  if (length(genes) < 1){
    paste("Select a gene.")
    }else{
      select(filter(Pval.annot, GeneName %in% genes), GeneName, adj.P.Val)
        }
  
  }, digits=10)

```



##20 most significant adj. P-values
Here we plot and describe the genes returning the 20 most significant adjusted P-values from t-test between the two groups. 
```{r, echo=FALSE}

renderPlot({
  #Get the 20 first rows of Pval.annot (contains the pvalues and gene info).
  #The df is order by adj.P.Val from lower to higher so the 20 first rows correspond to the 20 most significant pvalues
  Pval.annot<-Pval.annot()[1:20,]
  #get the names of the 20 most significant genes
  genes<-as.character(Pval.annot$GeneName)
  data.plot<-data.plot()
  #Remove empty string from the top 20 genes
  genes<-genes[genes != ""]  
  data.plot <- filter(data.plot, GeneName %in% genes)
  
  p<-ggplot(data.plot, aes(as.factor(SampleID), mean)) + theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.text = element_text(size = 12))
  p + geom_bar(aes(fill=group), stat="identity") +
    #   p + geom_bar(stat="identity") +
    geom_errorbar(aes(ymin= mean - sd, ymax=mean + sd), width=0.25) +
    facet_wrap(~GeneName)
  
  })

```

###Gene details
```{r, echo=FALSE, message=FALSE}
#Print a table with the gene details for the most 20 significant genes
renderTable({
  Pval.annot<-Pval.annot()[1:20,]
  t<-select(Pval.annot, Name, GeneName, Description, adj.P.Val)
  t
  
  }, digits=10)
```



##Significant adjusted P-values

Volcano plot showing the significance of the adj. P-values (cutoff at 0.01). Blue dots shows logFC returning significant P-values. 
```{r, echo=FALSE}

renderText({
  Pval.annot<-Pval.annot()
  nm<-as.character(length(which(Pval.annot$adj.P.Val < 0.01)))
  #   print("sdfs")
  paste("Number of pvalues < 0.01: ", nm)
  })


renderPlot({
  Pval.annot<-Pval.annot()
  #Create a column containing a logical vector of those adj.P.vals < 0.01
  Pval.annot$threshold = as.factor(Pval.annot$adj.P.Val < 0.01)
  
  #The volcano plot is extracted from here:
  # http://bioinformatics.knowledgeblog.org/2011/06/21/volcano-plots-of-microarray-data/
  
  g = ggplot(data=Pval.annot, aes(x=logFC, y=-log10(adj.P.Val), colour=threshold)) +
    geom_point(alpha=0.4, size=1.75) +
    theme(legend.position = "none") +
    scale_x_continuous(breaks=-10:10) +
    scale_y_continuous(breaks=0:10) +
    #   geom_text(aes(label=GeneName, size=1.2), colour="black") +
    #   xlim(c(-10, 10)) + ylim(c(0, 15)) +
    xlab("log2 fold change") + ylab("-log10 p-value")
  g
  })


```



